# Perturbation vector theory {#Chapter-Perturbation-vector}

## Objective

***
The objective of this chapter is to develop a mathematical workflow useful to adjust the ionome of potato crops for diagnostic purpose. Perturbation in compositional space plays the same role as translation plays in real space. Some natural processes in nature can be interpreted as a change from one composition `C1` to another `C2` through the application of a perturbation:

> `p ⊕ C1 ====> C2`.

The difference between `the new observation`  and the closest TN composition can be back-transformed to the compositional space. The obtained vector is the `perturbation vector`. Theoretically, a misbalanced composition could be balanced (translated into a healthy zone) using a perturbation vector. Also, ionome of a new cultivar could be assigned to the closest healthy ionomics group where nutrient requirements have been already documented by fertilizer trials.

I compute a dissimilarity index using the true negative clr as reference. Then, I describe the procedure used to rebalance a misbalanced composition and finally present the results using ggplot2 functions.

***

## Data set and useful libraries

We need package [compositions](https://www.rdocumentation.org/packages/compositions/versions/1.40-2) for further clr back-transformation de compositional space. The package [reshape](https://cran.r-project.org/web/packages/reshape/index.html) will be used to melt an intermediate data frame. As explained at the end of the Chapter \@ref(Chapter-Modeling), I consider as True Negatives (TN) specimens (or healthy points) for this study, observations of the `training data set` having a high yield (HY) and correctly predicted by the random forest model.

```{r, warning=FALSE, message=FALSE}
library("tidyverse")
library("extrafont")
library('compositions')
library("reshape")
train_df = read_csv("output/train_df.csv")
train_df$group_i = factor(train_df$group_i)
TNs = train_df[train_df$yieldClass == 'HY' & train_df$pred_yield == 'HY', ]
clrNo = c("clrN", "clrP", "clrK", "clrCa", "clrMg", "clrFv") # for simplicity
```

## Dissimilarity index between compositions

The first step is to compute a dissimilarity index according to the distance from the closest healthy point (the closest TN). I use the Euclidean distance between clr composition as the dissimilarity index. The chunk below shows the custom function used to compute euclidean distance.

```{r}
eucl_dist_f <- function(x, y) {
    sqrt(sum((x-y)^2))
}
```

For each unbalanced composition, I use the next loop to compute all the euclidean distances between all the compositions in "TNs" of the corresponding group. The computation is possible even if the ionomics groups is unknown. The loop returns the smallest Euclidean distance as the unbalanced index of that composition. The computed debalance (or imbalance) index is stored by a new vector `debal`.

```{r}
debal <- c()
debal_index <- c()
for (i in 1:nrow(train_df)) {
    clr_i <- as.numeric(train_df[i, clrNo])
    eucl_dist <- apply(TNs %>% filter(group_i == train_df$group_i[i]) %>% select(clrNo), 
                       1, function(x) eucl_dist_f(x=x, y=clr_i))
    debal_index[i] <- which.min(eucl_dist)
    debal[i] <- eucl_dist[debal_index[i]]
}
train_df$debal <- debal
train_df %>% glimpse()
```

## Rebalancing a misbalanced sample by perturbation

Let's suppose that we got this point selected at random in `unbalanced specimens`.

```{r}
set.seed(932559) 
unbalanced <- subset(train_df, debal !=0)
unbalanced <- unbalanced[sample(nrow(unbalanced), 1), ]
t(unbalanced)
```

Or even, we could rather use the most unbalanced occurrence, why not !

```{r}
unbalanced <- subset(train_df, debal !=0)
unbalanced <- unbalanced[which.max(unbalanced$debal), ]
misbalanced <- unbalanced # copy
t(misbalanced)
```

How could we rebalance it? The first step is to find `in TNs of the corresponding ionomic group` the closest balanced point. Let's re-compute its Euclidean distances from TNs and identify the sample from which the distance is minimum.

```{r}
misbalanced <- misbalanced[clrNo]
eucl_dist_misbal <- apply(TNs[, clrNo], 1, function(x) eucl_dist_f(x=x, y=misbalanced))
index_misbal <- which.min(t(data.frame(eucl_dist_misbal))) 
index_misbal # return the index of the sample
```

The closest healthy sample is the one which index is `r index_misbal`. Using this index we could refind the `debal` value of the misbalanced sample for check.

```{r}
(misbal = eucl_dist_misbal[index_misbal])
```

The Euclidean distance matches with the corresponding `debal` value: `unbalanced$debal[1]` = `r unbalanced$debal[1]`. 

The `closest point in the TNs` subset is this one:

```{r}
closest <- TNs[index_misbal, ]
t(closest)
```

Note that ionomics groups of the misbalanced and the closest healthy composition are the same. I compute the clr difference between the closest and the misbalanced points. 

```{r}
closest = closest[clrNo]
clr_diff = closest - misbalanced
t(clr_diff)
```

The perturbation vector is that clr difference back-transformed to leaf compositional space.

```{r}
comp_names <- c("N", "P", "K", "Ca", "Mg", "Fv")
perturbation_vector <- clrInv(clr_diff)
names(perturbation_vector) <- comp_names
t(perturbation_vector)
```

Next, we should compute the compositions of the clr coordinates of the misbalanced point, as well as the closest TN point. The vectors could be gathered in a table made up of perturbation vector, misbalanced composition and the closest reference sample (pmc).

```{r}
misbal_comp <- clrInv(misbalanced)
names(misbal_comp) <- comp_names

closest_comp <- clrInv(closest)
names(closest_comp) <- comp_names

pmc = rbind(perturbation_vector, misbal_comp, closest_comp)
rownames(pmc) = c("perturbation_vector","misbal_comp","closest_comp")
pmc
```

We could even check that the simplex is closed to 1 for each vector.

```{r}
sum(perturbation_vector); sum(misbal_comp); sum(closest_comp)
```

The closest composition minus the misbalanced composition should return the perturbation vector.

```{r}
print(closest_comp - misbal_comp) # soustraction
print(perturbation_vector)        # for comparison
```

Or even, perturb the misbalanced point by the perturbation vector, you should obtain the closest TN point:

```{r}
print(misbal_comp + perturbation_vector) # perturbation
print(closest_comp)                      # for comparison
```

So, the assumption is true. The next codes show the concept using plots. I aranged a data frame for ggplot.

```{r}
df = data.frame(rbind(misbalanced, closest, clr_diff))
vectors = c("observation", "reference", "perturbation")
df$vectors = factor(vectors)
dfreshape = melt(df) # reshapes df for ggplot
dfreshape
```

The next two chunks plot with dots and histograms for each vector. Visualization is better with histograms.

```{r hi-perturb-dotplot, fig.height = 3, fig.width = 6, out.width="100%", fig.align="center", warning=FALSE, message=FALSE, fig.cap="Perturbation vector computation example dotplot using the most imbalanced foliar sample."}
ggplot(data = dfreshape, aes(x = value, y = vectors)) +
    geom_point() +
    facet_wrap(~ variable, scales = "free_x") +
    labs(x='clr coordinate', y ='') +
    theme(text=element_text(family="Arial", face="bold", size=12))
#ggsave("images/perturb_dotplot.tiff",  width = 6, height = 3)
```



```{r hi-perturb-barplot, fig.height = 3, fig.width = 6, out.width="100%", fig.align="center", warning=FALSE, message=FALSE, fig.cap="Perturbation vector computation example barplot using the most imbalanced foliar sample."}
ggplot(data=dfreshape, aes(x=variable, y=value, fill=vectors)) +
    geom_bar(stat="identity", position=position_dodge()) +
    coord_flip() +
    theme_bw() +
    theme(legend.title=element_blank()) +
    theme(text=element_text(family="Arial", face="bold", size=12))
#ggsave("images/perturb_barplot.tiff",  width = 6, height = 3)
```

