# Perturbation vector theory

The first step is to compute a dissimilarity index according to the distance from the closest healthy point (the closest TN). Let's load data set.

```{r, warning=FALSE, message=FALSE}
library("tidyverse")     # dplyr and ggplot2
library(extrafont)        # Changing Fonts for Graphs

train_df = read.csv2("data/train_df.csv")
train_df <- train_df %>% select(-X.1, -X) # to discard extra columns
train_df$group_i = factor(train_df$group_i)
```

As considered in previous `1.2_machine learning.ipynb` file, True Negatives (TN) specimens in this study are observations of the training data set with high yield (HY) and correctly predicted.

```{r}
TNs = train_df[train_df$yieldClass == 'HY' & train_df$pred_yield == 'HY', ]
clrNo = c("clrN", "clrP", "clrK", "clrCa", "clrMg", "clrFv") # for simplicity
```

### Dissimilarity index between compositions

- Euclidean distance as dissimilarity index
- For each composition (line) in the "unbalanced specimens", calculate all the euclidean distances between all the compositions in "TNs" of the corresponding group
- Return the smallest euclidean distance as the unbalanced index of the composition.

Function to compute euclidean distance

```{r}
eucl_dist_f <- function(x, y) {
    sqrt(sum((x-y)^2))
}
```

Compute debalance (or imbalance) index (`debal`) using this loop.

```{r}
debal <- c()
debal_index <- c()
for (i in 1:nrow(train_df)) {
    clr_i <- as.numeric(train_df[i, clrNo])
    eucl_dist <- apply(TNs %>% filter(group_i == train_df[i, 'group_i']) %>% select(clrNo), 
                       1, function(x) eucl_dist_f(x=x, y=clr_i))
    debal_index[i] <- which.min(eucl_dist)
    debal[i] <- eucl_dist[debal_index[i]]
}
train_df$debal <- debal
train_df %>% glimpse()
```

### Rebalancing a misbalanced sample by perturbation

Suppose we got this point selected at random in `unbalanced specimens`.

```{r}
#set.seed(932559) # random.org
unbalanced <- subset(train_df, debal !=0)
misbalanced <- unbalanced[sample(nrow(unbalanced), 1), ]
misbalanced
```

Or even, you can use the most unbalanced occurrence, ... why not !

```{r}
(misbalanced = unbalanced[which.max(unbalanced$debal), ])
```

How to rebalance it?

The first step is to find `in TNs of the corresponding ionomic group` the closest balanced point.

Let's re-compute the euclidean distance, like for a new (an unknown) point:

```{r}
misbalanced <- misbalanced[clrNo]
eucl_dist_misbal <- apply(TNs[, clrNo], 1, function(x) eucl_dist_f(x=x, y=misbalanced))
index_misbal <- which.min(t(data.frame(eucl_dist_misbal))) # return the index of the sample
index_misbal
```

Euclidean distance matched with the corresponding `debal` value:

```{r}
misbal = eucl_dist_misbal[index_misbal] # to compare to corresponding debal value
misbal
```

The `closest point in the TNs` data set is this one:

```{r}
closest <- TNs[index_misbal, ]
closest
```

Note that Ionomics groups of the misbalanced and the closest composition are the same ... most of the times ! You need package `compositions` for further clr back-transformation de compositional space.

```{r, warning=FALSE, message=FALSE}
require('compositions')
```

Perturbation in compositional space plays the same role as translation plays in real space. Some natural processes in nature can be interpreted as a change from one composition `C1` to another `C2` through the application of a perturbation:

> `p ⊕ C1 ====> C2`.

The difference between `the new observation`  and the closest TN composition can be back-transformed to the compositional space. The obtained vector is the `perturbation vector`.

```{r}
closest = closest[clrNo]
clr_diff = closest - misbalanced
clr_diff
```

```{r}
comp_names <- c("N", "P", "K", "Ca", "Mg", "Fv")
perturbation_vector <- clrInv(clr_diff)
names(perturbation_vector) <- comp_names
```

Compute the compositions of the ilr coordinates of the misbalanced point, as well as the closest TN point:

```{r}
misbal_comp <- clrInv(misbalanced)
names(misbal_comp) <- comp_names

closest_comp <- clrInv(closest)
names(closest_comp) <- comp_names

pmc = rbind(perturbation_vector, misbal_comp, closest_comp)
rownames(pmc) = c("perturbation_vector","misbal_comp","closest_comp")
pmc # data frame made up of perturbation vector, misbalanced composition and the closest reference sample
```

For each vector, check that the simplex is closed to 1.

```{r}
sum(perturbation_vector); sum(misbal_comp); sum(closest_comp)
```

The closest composition minus the misbalanced composition should return the perturbation vector.

```{r}
print(closest_comp - misbal_comp) # soustraction
print(perturbation_vector)        # for comparison
```

Or even, perturb the misbalanced point by the perturbation vector, you should obtain the closest TN point:

```{r}
print(misbal_comp + perturbation_vector) # perturbation
print(closest_comp)                      # for comparison
```

So, the assumption is true.

The next codes try to show the concept using plots using clr coordinates. 

```{r}
d = data.frame(rbind(misbalanced, closest, clr_diff)) # mis, cl, per
vectors = c("observation", "reference", "perturbation")
d$vectors = factor(vectors)
d
```

This data frame must be reshape for ggplot.

```{r, warning=FALSE, message=FALSE}
require("reshape") # function melt() 
dreshape = melt(d)
dreshape
```

Plot with dots for each vector.

```{r hi-perturb-dotplot, out.width="100%", fig.align="center", warning=FALSE, message=FALSE, fig.cap="Perturbation vector computation example dotplot using the most imbalanced foliar sample."}
options(repr.plot.width = 6, repr.plot.height = 3)
ggplot(data = dreshape, aes(x = value, y = vectors)) +
    geom_point() +
    facet_wrap(~ variable, scales = "free_x") +
    labs(x='clr coordinate', y ='') +
    theme(text=element_text(family="Arial", face="bold", size=12))
#ggsave("images/perturb_dotplot.tiff",  width = 6, height = 4)
```



```{r hi-perturb-barplot, out.width="100%", fig.align="center", warning=FALSE, message=FALSE, fig.cap="Perturbation vector computation example barplot using the most imbalanced foliar sample."}
options(repr.plot.width = 5, repr.plot.height = 3)
ggplot(data=dreshape, aes(x=variable, y=value, fill=vectors)) +
    geom_bar(stat="identity", position=position_dodge()) +
    coord_flip() +
    theme_bw() +
    theme(legend.title=element_blank()) +
    theme(text=element_text(family="Arial", face="bold", size=12))
ggsave("images/perturb_barplot.tiff",  width = 6, height = 4)
```

