--- 
title: "Nutrient Diagnosis Of Potato"
author: "zcoulibali"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
output: bookdown::gitbook
description: "This is a example of using the bookdown package to write a book for publication on Github. The output format for this example is bookdown::gitbook. Next features are set for after (bibliography and links)."
#bibliography: [book.bib, packages.bib]
#biblio-style: apalike
#link-citations: yes
---

# Data processing {#Chapter-Data-Processing}
## Data sets

We need package `tidyverse` for data handling, `DBI` and `RSQLite` packages to make connexion and extract usefull tables in the potato historical database file `pomme_de_terre.db`.

```{r, message=FALSE, warning=FALSE}
library("tidyverse")  # loads dplyr & ggplot2
library("DBI")        # Database Interface for R
library("RSQLite")    # SQLite Interface for R
db <- dbConnect(SQLite(), dbname = "data/pomme_de_terre.db")  # Connect to potato database
```

Select tables in the created connexion.

```{r, message=FALSE, warning=FALSE}
data_df <- dbReadTable(db, "MetaData") %>% 
  left_join(dbReadTable(db, "FoliarAnalysis"), by='NoEssai') %>% 
  left_join(dbReadTable(db, "TreatmentVariable"), by=c('NoEssai', 'NoBloc', 'NoTraitement')) %>% 
  left_join(dbReadTable(db, "MaturityOrder"), by = 'Cultivar') %>% 
  as_tibble()
```

Load recent collected trials data from the project experiments and the Quebec Ministry of Agriculture, Fisheries and Food (MAPAQ) trials.

```{r}
trials_df <- read.csv2("data/donneesEssaisPdt.csv", sep = ';', dec = '.')
```

Select usefull variables

Select usefull columns for computations, with macroelements `N, P, K, Mg, Ca` only because oligoelements have too much missing data. The year is not needed instead it permits to know how long ago expériements have been monitored. Geographical coordinates are used to map sites locations.

```{r}
keys_col <- c('NoEssai', 'NoBloc', 'NoTraitement')
macro <- c("AnalyseFoliaireN","AnalyseFoliaireP","AnalyseFoliaireK", "AnalyseFoliaireCa","AnalyseFoliaireMg")
coord_year <- c('Annee', 'LatDD', 'LonDD')
cult_yield <- c('Cultivar', 'Maturity5', 'RendVendable')
usefull_col <- c(keys_col, macro, coord_year, cult_yield, 'AnalyseFoliaireStade')
macro_elts <- c("N", "P", "K", "Ca", "Mg") # for simplicity
```

Reduced and joined data frame becomes `fol_df` the foliar data frame:

```{r}
data_df <- data_df %>% select(usefull_col)
trials_df <- trials_df %>% select(usefull_col)
fol_df <- rbind(data_df, trials_df)
colnames(fol_df)[which(names(fol_df) %in% macro)] <- macro_elts
glimpse(fol_df)
```

## Arranging the data frame

Set trial number as factor

```{r}
fol_df$NoEssai <- as.factor(fol_df$NoEssai)
```

Choose cultivar `Goldrush` as reference, it as the maximum number of observations in the data frame.

```{r}
fol_df$Cultivar <-  relevel(factor(fol_df$Cultivar), ref = "Goldrush")
```

Relevel categorical `maturity` order.

```{r}
fol_df$Maturity5 <- ordered(fol_df$Maturity5, 
                            levels = c("early","early mid-season","mid-season","mid-season late","late"))
```

Leaf ionome with macroelements

```{r}
leafIonome <- fol_df[macro_elts]
```

Some custom functions are used for compositional analysis. The libraries `robCompositions`, `compositions` and `Amelia` are used for robust imputation with kNN (long process), compositional transformations and to portrait missing values respectively.

```{r}
source('https://raw.githubusercontent.com/essicolo/AgFun/master/ilrNA.R')
source('https://raw.githubusercontent.com/essicolo/AgFun/master/ilrDefinition.R')
source("https://raw.githubusercontent.com/essicolo/AgFun/master/codadend2.R")
```

```{r, message=FALSE, warning=FALSE}
library("robCompositions")  # impCoda & impKNNa: for data imputation
library("compositions")  # for ILR transformations: acomp, ilr, ilrInv
require('Amelia')   # Portrait of missing values
```

Portrait of missing macroelements

```{r matrix-missing-data, out.width="100%", fig.align="center", warning=FALSE, message=FALSE, fig.cap="Portrait of missing macroelements."}
missmap(leafIonome)
```

Impute missing data for samples (rows) with less than 3 missing elements among the five. The next cell initializes this computation codes.

```{r}
# keep track of empty rows:
fol_df$leafIonome_allNA <- apply(fol_df[macro_elts], 1, function(X) all(is.na(X)))
# keep track of rows where there is any NA:
fol_df$leafIonome_anyNA <- apply(fol_df[macro_elts], 1, function(X) any(is.na(X)))
# number of NAs (missing values):
fol_df$leafIonome_countNA <- apply(fol_df[macro_elts], 1, function(X) sum(is.na(X)))
# Only impute if the next variable value is set to FALSE
fol_df$leafIonome_hasTooMuchNA <- fol_df$leafIonome_countNA >= 3
```

Imputation:

```{r}
# Warning: could be a long process
leaf_imputeRob <- impKNNa(as.matrix(fol_df[!fol_df$leafIonome_hasTooMuchNA, macro_elts]),
                           metric = "Aitchison", k = 6, primitive = TRUE,
                           normknn = TRUE, adj = 'median')
colnames(leaf_imputeRob$xImp) <- paste0(colnames(leaf_imputeRob$xImp), '_imp')
leaf_imputeRob$xImp %>% head() # a view of new imputed compositions
```

Push imputed columns to the data frame. The nutrients diagnosis will be performed with imputed compositions.

```{r}
fol_df <- left_join(x = fol_df,
                     y = data.frame(NoEssai = fol_df$NoEssai[!fol_df$leafIonome_hasTooMuchNA],
                                    NoBloc = fol_df$NoBloc[!fol_df$leafIonome_hasTooMuchNA],
                                    NoTraitement = fol_df$NoTraitement[!fol_df$leafIonome_hasTooMuchNA],
                                    leaf_imputeRob$xImp),
                     by = keys_col)
fol_df <- fol_df %>% select(-c("leafIonome_allNA", "leafIonome_anyNA", "leafIonome_countNA",                                                     "leafIonome_hasTooMuchNA"))
```

Compute Fv. `Fv` stands for filling value, an amalgamation of all other elements closing the simplex to 100%.

```{r}
fol_df$Fv_imp <- 100 - rowSums(fol_df[, colnames(leaf_imputeRob$xImp)]) # computes the Fv
leaf.macro_imp <- c(colnames(leaf_imputeRob$xImp), 'Fv_imp') # new colnames vector of macroelements
leaf.macro_imp
```

Compute centered log-ratios `clr` only for complete cases.

```{r}
leafIonomeComp <- acomp(fol_df[leaf.macro_imp])
leafIonomeClr <- clr(leafIonomeComp)
```

```{r}
leafIonomeClr[apply(leafIonomeComp, 1, anyNA), ] <- NA   # discard for rows with any NA
leafIonomeDefClr <- paste0("clr_", c(macro_elts, 'Fv'))  # computed clr colnames vector
colnames(leafIonomeClr) <- leafIonomeDefClr
leafIonomeClr <- cbind(fol_df[keys_col], leafIonomeClr)  # bind clr variables to keys columns
```

Push clr coordinates to `fol_df` and reduce the data frame to usefull columns

```{r}
fol_df <- left_join(fol_df, leafIonomeClr, by = keys_col)
fol_df <- fol_df %>% select(keys_col, coord_year, cult_yield, "AnalyseFoliaireStade", leafIonomeDefClr)
glimpse(fol_df)
```

Cultivars classes correction. 

Cultivar `Mystere` and `Vivaldi` have different maturity classes in the data set, `mid-season late` and `late` for `Mystere`, then `early mid-season` and `mid-season` for `Vivaldi`. Their new maturity classes names are based on a majority vote.

```{r}
fol_df$Maturity5[fol_df$Cultivar == "Mystere"] <- "late"
fol_df$Maturity5[fol_df$Cultivar == "Vivaldi"] <- "early mid-season"
fol_df$Cultivar <- forcats::fct_explicit_na(fol_df$Cultivar)
```

Summarise the data frame.

```{r}
fol_df %>%
  summarise(start_year = min(Annee, na.rm = TRUE),
            end_year = max(Annee, na.rm = TRUE),
            nbr_trials = n_distinct(NoEssai, na.rm = TRUE),
            nbr_cultivars = n_distinct(Cultivar, na.rm = TRUE),
            nbr_maturityClass = n_distinct(Maturity5, na.rm = TRUE)
            )
```

Backup for cluster analysis:

```{r}
write.csv2(fol_df, 'data/fol_df.csv')
```

## Map of experimental sites locations

```{r fol-locations, out.width="100%", fig.align="center", warning=FALSE, message=FALSE, fig.cap="Location of experimental sites (green dots) in the Québec potato data set."}
library("ggmap") # maps with ggplot2
library("extrafont") # Changing Fonts for Graphs
qc_fol <- get_stamenmap(bbox = c(left=-76, right=-68, bottom=45, top=50), 
                        zoom=7, maptype = 'toner-lite')
ggmap(qc_fol) +
  geom_point(data=unique(fol_df[c('LonDD', 'LatDD')]), aes(x=LonDD, y=LatDD),
             size=3, shape=16, colour='aquamarine4', alpha=0.8) + coord_map("mercator") +
  theme_bw() +
  theme(text=element_text(family="Arial", face="bold", size=12))
#ggsave("images/fol-locations.png", width=10, height=8)
```
